---
title: "Análisis longitudinal: RI-CLPM"

author: Equipo EDUMER
subtitle: SoJ
fontsize: 12pt
#bibliography: "../input/mini-proyecto.bib"
#csl: "input/bib/apa6.csl"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 6
    toc-expand: 6
    toc-title: Contenidos
    number-sections: false
    number-depth: 6
    theme: 
    - edumer.scss
    - sandstone
    title-block-banner: true
editor: visual
lang: es
---


```{r,data,  echo=FALSE}
pacman::p_load(dplyr, sjmisc, car, sjlabelled, stargazer, haven, sjPlot, ggplot2, tidyr, stringr, corrplot, lavaan, data.table)

load("../input/data/proc/edumer_students_long.RData")
setDT(edumer_students_long)
#edumer_students_long <- edumer_students_long[muestra==1,]

```

```{r, preparacion, echo=FALSE}

# recodificar na 
edumer_students_long <- edumer_students_long[, lapply(.SD, function(x) replace(x, which(x %in% c(88,99)), NA))]

# indice de justicia 
# nota preferida 
edumer_students_long <- edumer_students_long %>%
  mutate(nota_just = case_when(
    is.na(p6) & p5 == 2 ~ p4 ,  # Imputar el valor de 'notas_merit' (2) cuando 'nota_pref' es NA
    TRUE ~ p6  # Mantener los valores existentes de 'nota_pref' cuando no son NA
  ))

edumer_students_long <- edumer_students_long %>% mutate(just_school = log(p4/nota_just))
                                    
#summary(edumer_students_long$just_school)

# long to wide
datos <- dcast(edumer_students_long, id_estudiante~ola,
               value.var = c("just_school", "p1_1", "p1_2", "p1_3", "p2_1", "p2_2", "p2_3", "p30", "p26", "p27", "d3_def", "p20" ))

# seleccionar variables
old <- c("id_estudiante", paste0("just_school_", 1:2), paste0("p1_1_", 1:2), paste0("p1_2_", 1:2), paste0("p1_3_", 1:2), paste0("p2_1_", 1:2), paste0("p2_2_", 1:2), paste0("p2_3_", 1:2), "p30_1", "p26_1", "p27_1", "d3_def_1", "p20_1")
new <- c("id", paste0("just_school", 1:2), paste0("merit_se", 1:2),paste0("merit_st", 1:2),paste0("merit_soc", 1:2), paste0("merit_sce", 1:2),paste0("merit_sct", 1:2),paste0("merit_school", 1:2),"libros_hogar", "ne_madre", "ne_padre", "curso", "genero")
setnames(datos,old,new)
datos <- datos[, ..new]

# cleanning workspace 
datos <- as.data.frame(datos)

```

# Modelo Básico 

```{r}

```

# Ambos (Extensión RI-CLPM con predictores)

Solo con las variables de percepción de meritocracia

## Paso 1: Modelo configuracional

```{r, echo=FALSE}
RICLPM1.ext3 <- '
  
  ################
  # BETWEEN PART #
  ################
  
  # Create between factors (random intercepts) for each indicator separately
  
  RIX1 =~ 1*merit_se1 + 1*merit_se2 
  RIX2 =~ 1*merit_st1 + 1*merit_st2
  RIX3 =~ 1*merit_soc1 + 1*merit_soc2
  
  RIY1 =~ 1*merit_sce1 + 1*merit_sce2 
  RIY2 =~ 1*merit_sct1 + 1*merit_sct2
  RIY3 =~ 1*merit_school1 + 1*merit_school2
  
  
  ##################################
  # WITHIN PART: MEASUREMENT MODEL #
  ##################################
  
  # Factor models for Y at 5 waves
  WFX1 =~ merit_se1 + merit_st1 + merit_soc1
  WFX2 =~ merit_se2 + merit_st2 + merit_soc2

  # Factor models for Z at 5 waves
  WFY1 =~ merit_sce1 + merit_sct1 + merit_school1
  WFY2 =~ merit_sce2 + merit_sct2 + merit_school2
  
  #########################
  # WITHIN PART: DYNAMICS #
  #########################
  
  # Specify lagged effects between within-person centered latent variables
  WFX2 + WFY2 ~ WFX1 + WFY1
  
  # Estimate correlations within same wave
  WFX1 ~~ WFY1
  WFX2 ~~ WFY2
  
  ##########################
  # ADDITIONAL CONSTRAINTS #
  ##########################
  
  # Constrain covariance of between factors and exogenous within factors to 0 
  RIX1 + RIX2 + RIX3 + RIY1 + RIY2 + RIY3 ~~ 0*WFY1 + 0*WFX1
'
RICLPM1.ext3.fit <- cfa(RICLPM1.ext3, 
  data = datos, 
  missing = 'ML',
  meanstructure = TRUE
)
summary(RICLPM1.ext3.fit, standardized = T)
```

## Paso 2: Invarianza factorial débil

```{r, echo=FALSE}
RICLPM2.ext3 <- '
  
  ################
  # BETWEEN PART #
  ################
  
  # Create between factors (random intercepts) for each indicator separately
  RIX1 =~ 1*merit_se1 + 1*merit_se2 
  RIX2 =~ 1*merit_st1 + 1*merit_st2
  RIX3 =~ 1*merit_soc1 + 1*merit_soc2
  
  RIY1 =~ 1*merit_sce1 + 1*merit_sce2 
  RIY2 =~ 1*merit_sct1 + 1*merit_sct2
  RIY3 =~ 1*merit_school1 + 1*merit_school2
  
  ##################################
  # WITHIN PART: MEASUREMENT MODEL #
  ##################################
  
  # Factor models for X at 2 waves (constrained)
  WFX1 =~ a*merit_se1 + b*merit_st1 + c*merit_soc1
  WFX2 =~ a*merit_se2 + b*merit_st2 + c*merit_soc2
 
  # Factor models for Y at 2 waves (constrained)
  WFY1 =~ d*merit_sce1 + e*merit_sct1 + f*merit_school1
  WFY2 =~ d*merit_sce2 + e*merit_sct2 + f*merit_school2
 
  
  #########################
  # WITHIN PART: DYNAMICS #
  #########################
  
  # Specify lagged effects between within-person centered latent variables
  WFX2 + WFY2 ~ WFX1 + WFY1
  
  # Estimate correlations within same wave
  WFX1 ~~ WFY1
  WFX2 ~~ WFY2
  
  ##########################
  # ADDITIONAL CONSTRAINTS #
  ##########################
  
  # Constrain covariance of between factors and exogenous within factors to 0
  RIX1 + RIX2 + RIX3 + RIY1 + RIY2 + RIY3 ~~ 0*WFY1 + 0*WFX1
'
RICLPM2.ext3.fit <- cfa(RICLPM2.ext3, 
  data = datos, 
  missing = 'ML'
)
summary(RICLPM2.ext3.fit, standardized = T)
```

## Paso 3: Fuerte invarianza factorial

```{r, echo=FALSE}
RICLPM3.ext3 <- '
  
  ################
  # BETWEEN PART #
  ################
  
  # Create between factors (random intercepts) for each indicator separately
  RIX1 =~ 1*merit_se1 + 1*merit_se2 
  RIX2 =~ 1*merit_st1 + 1*merit_st2
  RIX3 =~ 1*merit_soc1 + 1*merit_soc2
  
  RIY1 =~ 1*merit_sce1 + 1*merit_sce2 
  RIY2 =~ 1*merit_sct1 + 1*merit_sct2
  RIY3 =~ 1*merit_school1 + 1*merit_school2
  
  ##################################
  # WITHIN PART: MEASUREMENT MODEL #
  ##################################
  
  # Factor models for X at 2 waves (constrained)
  WFX1 =~ a*merit_se1 + b*merit_st1 + c*merit_soc1
  WFX2 =~ a*merit_se2 + b*merit_st2 + c*merit_soc2
 
  # Factor models for Y at 2 waves (constrained)
  WFY1 =~ d*merit_sce1 + e*merit_sct1 + f*merit_school1
  WFY2 =~ d*merit_sce2 + e*merit_sct2 + f*merit_school2
  
  
  # Constrained intercepts over time (necessary for strong factorial 
  # invariance; without these contraints we have week factorial invariance) 
  merit_se1 + merit_se2 ~ g*1
  merit_st1 + merit_st2 ~ h*1
  merit_soc1 + merit_soc2 ~ i*1
  merit_sce1 + merit_sce2 ~ j*1
  merit_sct1 + merit_sct1 ~ k*1
  merit_school1 + merit_school2 ~ l*1
  
  # Free latent means from t = 2 onward (only do this in combination with  
  # constraints on intercepts; without these, this would not be specified)
  WFX2 + WFY2 ~ 1
  
  #########################
  # WITHIN PART: DYNAMICS #
  #########################
  
  # Specify lagged effects between within-person centered latent variables
  WFX2 + WFY2 ~ WFX1 + WFY1
  
  # Estimate correlations within same wave
  WFX1 ~~ WFY1
  WFX2 ~~ WFY2
  
  ##########################
  # ADDITIONAL CONSTRAINTS #
  ##########################
  
  # Constrain covariance of between factors and exogenous within factors to 0
  RIX1 + RIX2 + RIX3 + RIY1 + RIY2 + RIY3 ~~ 0*WFY1 + 0*WFX1
'
RICLPM3.ext3.fit <- cfa(RICLPM3.ext3, 
  data = datos, 
  missing = 'ML'
)
summary(RICLPM3.ext3.fit, standardized = T)
```

## Extra: Fuerte invarianza factorial con cargas factoriales iguales a las cargas factoriales dentro de la persona

```{r, echo=FALSE}
RICLPM4.ext3 <- '
  
  ################
  # BETWEEN PART #
  ################
  
  # Create between factors (random intercepts) for each indicator separately
  RIX1 =~ 1*merit_se1 + 1*merit_se2 
  RIX2 =~ 1*merit_st1 + 1*merit_st2
  RIX3 =~ 1*merit_soc1 + 1*merit_soc2
  
  RIY1 =~ 1*merit_sce1 + 1*merit_sce2 
  RIY2 =~ 1*merit_sct1 + 1*merit_sct2
  RIY3 =~ 1*merit_school1 + 1*merit_school2
  
  # Create a single random intercept for all X variables, and another for all 
  # Y variables and constrain the factor loadings to be identical to
  # within-person factor loadings
  
  RIX =~ a*RIX1 + b*RIX2 + c*RIX3
  RIY =~ d*RIY1 + e*RIY2 + f*RIY3
  
  # Add (co)variance between two higher-order random intercepts
  RIX ~~ RIY
  RIX ~~ RIX
  RIY ~~ RIY
  
  # Constrain measurement error variances of the second order factor model to 0 
  RIX1 ~~ 0*RIX1
  RIX2 ~~ 0*RIX2
  RIX3 ~~ 0*RIX3
  RIY1 ~~ 0*RIY1
  RIY2 ~~ 0*RIY2
  RIY3 ~~ 0*RIY3
  
  ##################################
  # WITHIN PART: MEASUREMENT MODEL #
  ##################################
  
 # Factor models for X at 2 waves (constrained)
  WFX1 =~ a*merit_se1 + b*merit_st1 + c*merit_soc1
  WFX2 =~ a*merit_se2 + b*merit_st2 + c*merit_soc2
 
  # Factor models for Y at 2 waves (constrained)
  WFY1 =~ d*merit_sce1 + e*merit_sct1 + f*merit_school1
  WFY2 =~ d*merit_sce2 + e*merit_sct2 + f*merit_school2
  
  # Constrained intercepts over time (necessary for strong factorial invariance; 
  # without these contraints we have week factorial invariance)
 merit_se1 + merit_se2 ~ g*1
  merit_st1 + merit_st2 ~ h*1
  merit_soc1 + merit_soc2 ~ i*1
  merit_sce1 + merit_sce2 ~ j*1
  merit_sct1 + merit_sct1 ~ k*1
  merit_school1 + merit_school2 ~ l*1
  
  # Free latent means from t = 2 onward (only do this in combination with  
  # constraints on intercepts; without these, this would not be identified)
  WFX2 + WFY2 ~ 1
  
  #########################
  # WITHIN PART: DYNAMICS #
  #########################
  
  # Specify lagged effects between within-person centered latent variables
  WFX2 + WFY2 ~ WFX1 + WFY1
  
  # Estimate correlations within same wave
  WFX1 ~~ WFY1
  WFX2 ~~ WFY2
  
  ##########################
  # ADDITIONAL CONSTRAINTS #
  ##########################
  
  # Constrain covariance of between factors and exogenous within factors to 0 
  RIX + RIY + RIX1 + RIX2 + RIX3 + RIY1 + RIY2 + RIY3 ~~ 0*WFY1 + 0*WFX1
'
RICLPM4.ext3.fit <- cfa(RICLPM4.ext3, 
  data = datos, 
  missing = 'ML'
)
summary(RICLPM4.ext3.fit, standardized = T)
```

## Paso 4: RI-CLPM latente

```{r, echo=FALSE}
RICLPM5.ext3 <- '
  
  #####################
  # MEASUREMENT MODEL #
  #####################
  
 # Factor models for X at 2 waves (constrained)
  FX1 =~ a*merit_se1 + b*merit_st1 + c*merit_soc1
  FX2 =~ a*merit_se2 + b*merit_st2 + c*merit_soc2
 
  # Factor models for Y at 2 waves (constrained)
  FY1 =~ d*merit_sce1 + e*merit_sct1 + f*merit_school1
  FY2 =~ d*merit_sce2 + e*merit_sct2 + f*merit_school2
  
  # Constrained intercepts over time (this is necessary for strong factorial 
  # invariance; without these contraints we have week factorial invariance)
 merit_se1 + merit_se2 ~ g*1
  merit_st1 + merit_st2 ~ h*1
  merit_soc1 + merit_soc2 ~ i*1
  merit_sce1 + merit_sce2 ~ j*1
  merit_sct1 + merit_sct1 ~ k*1
  merit_school1 + merit_school2 ~ l*1
  
  # Free latent means from t = 2 onward (only do this in combination with 
  # constraints on intercepts; without these, this would not be identified)
  FX2 + FY2 ~ 1
  
  ################
  # BETWEEN PART #
  ################
  
  # Create between factors (random intercepts)
  RIX =~ 1*FX1 + 1*FX2 
  RIY =~ 1*FY1 + 1*FY2 
  
  # Set residual variances of all FX and FY variables to 0
  FX1 ~~ 0*FX1
  FX2 ~~ 0*FX2
  FY1 ~~ 0*FY1
  FY2 ~~ 0*FY2
  
  ###############
  # WITHIN PART #
  ###############
 
  # Create within-components
  WFX1 =~ 1*FX1
  WFX2 =~ 1*FX2
  
  WFY1 =~ 1*FY1
  WFY2 =~ 1*FY2
  
  # Specify lagged effects between within-person centered latent variables
  WFX2 + WFY2 ~ WFX1 + WFY1
  
  # Estimate correlations within same wave
  WFX2 ~~ WFY2
  
  ##########################
  # ADDITIONAL CONSTRAINTS #
  ##########################
  
  # Set correlations between between-factors (random intercepts) and within-
  # factors at wave 1 at 0
  RIX + RIY ~~ 0*WFX1 + 0*WFY1
'
RICLPM5.ext3.fit <- cfa(RICLPM5.ext3,
  data = datos, 
  missing = 'ML'
)
summary(RICLPM5.ext3.fit, standardized = T)
```

# Ambos (Extensión RI-CLPM con controles)

```{r, echo=FALSE}
RICLPM.ext1 <- '
  # Create between components (random intercepts)
  
  RIx =~ 1*just_school1 + 1*just_school2 
   
  RIy =~ 1*merit_soc1 + 1*merit_soc2 
    
  RIz =~ 1*merit_school1 + 1*merit_school2 
  
  # Create within-person centered variables
  wx1 =~ 1*just_school1
  wx2 =~ 1*just_school2
  wy1 =~ 1*merit_soc1
  wy2 =~ 1*merit_soc2
  wz1 =~ 1*merit_school1
  wz2 =~ 1*merit_school2
  
  # Regression of observed variables on z1 (constrained)
  just_school1 + just_school2 ~ s1*genero # Constrained over time. 
  merit_soc1 + merit_soc2 ~ s2*genero # Constrained over time. 
  merit_school1 + merit_school2 ~ s3*genero
  

  # Estimate lagged effects between within-person centered variables
  wx2 + wy2 ~ wx1 + wy1
  wx2 + wz2 ~ wx1 + wz1
  wz2 + wy2 ~ wz1 + wy1
  
  # Estimate covariance between within-person centered variables at first wave
  wx1 ~~ wy1 # Covariance
  wx1 ~~ wz1
  wy1 ~~ wz1
  
  # Estimate covariances between residuals of within-person centered variables
  # (i.e., innovations)
  wx2 ~~ wy2
  wx2 ~~ wz2
  wy2 ~~ wz2
 
  
  # Estimate variance and covariance of random intercepts
  RIx ~~ RIx
  RIy ~~ RIy
  RIz ~~ RIz
  RIx ~~ RIy
  RIz ~~ RIx
  RIz ~~ RIy
  
  # Estimate (residual) variance of within-person centered variables
  wx1 ~~ wx1 # Variances
  wy1 ~~ wy1 
  wx2 ~~ wx2 # Residual variances
  wy2 ~~ wy2 
  wz1 ~~ wz1
  wz2 ~~ wz2
'
RICLPM.ext1.fit <- lavaan(RICLPM.ext1, 
  data = datos, 
  missing = 'ML', 
  meanstructure = T, 
  int.ov.free = T
) 
summary(RICLPM.ext1.fit, standardized = T)
```
